Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-02-01 paper/rebuttal/rebuttal.tex

We have also provided detail about how confidence intervals were calculated in relevant figure captions. 
For example, we state in figure 2's caption that \say{\textit{99\% confidence intervals are plotted for the mean flux in each bin based on a normal distribution.}}, which should give readers enough information to reproduce the confidence intervals following their standard definition. 
Other confidence intervals, such as intervals for the loess regression line in figure 6, are calculated using the smoothing functions in the \texttt{ggplot2} package of R, and refer readers to the relevant documentation. 
}

\rcomment{%
F. Suggested improvements:
References and discussion of prior work using word2vec-style embeddings of mobility data.
Additional experiments using alternative embedding methods, especially an SVD-based embedding method, as well as robustness/reproducibility studies across different runs of the SGNS would be essential to strengthen the paper.
In addition, some additional results demonstrating how the embeddings provide novel scientific insight, rather than just recapitulating known structures, would be essential to make this work above-the-bar for a high-impact publication such as Nature.

}

\response{
We believe that the mathematical breakthroughs made since our first submission improve the manuscript and extend its contribution and demonstrates novel scientific insight, and justifies the use of word2vec towards mobility data. By adding this, and addressing your other points (above), we believe that the revised manuscript sufficiently remedies your concerns.

@@

\rcomment{%
Although the paper is well written and explores a very interesting problem, I have certain criticism for the used methodology. After obtaining a representation of each dataset, the authors defined an embedding distance as the cosine distance between vectors in the space, while defining distance as the geographical distance in the original, spatially embedded datasets. Subsequently they used the Gravity Law to estimate the flux between location pairs of vector pairs in the spatial and embedded space respectively. The Gravity Law, applied on mobility, assumes that the number of people moving between two populations is proportional to the multiplicative of the sizes of the populations and a deterrence function. Research during the last decades on transportation, spatial networks and human mobility suggests that this deterrence function may take a functional form of an inverse exponential or an inverse power-law, as it has been acknowledged by the authors. However, the emergence of these different deterrence functions can be due to fundamentally different mechanisms (for a review on the topic see Barthélemy, Physics Reports 499, 1, 2011, a paper not cited in the manuscript). For example, the different cost functions (linear or logarithmic) to create link at a given distance, the spatial granularity of the observation, or even the heterogeneity of the moving individuals in terms of various factors, just to mention a few examples. Instead, the authors of the actual paper arbitrarily assumed a power-law deterrence function in the spatial case and an exponential deterrence function in the embedding space without further arguing their choice (other than fitting on the data). Furthermore, assuming different deterrence functions for the flux estimation using Gravity Law makes the estimated fluxes incomparable, resulting the invalidation of results depicted in Figure 2.
}

\response{%
Thank you for pointing this out. 
We carefully read out the Barthélemy, Physics Reports 499, 1, 2011, and referenced it in our manuscript. 
Also, we report the performance of the flux estimation (Fig.~\ref{fig:supp:distancemetrics}, this document) and prediction (Fig.~\ref{fig:supp:distancermse}, this document) using different deterrence functions, funding that while performance differs, that the embedding distance outperforms geographic 
distance no matter the function used. 
We add the following line to the methods,

\medskip
\say{The decay function $f(r_{ij})$ defines the effect of distance, and so different decay functions can model fundamentally different mechanism [...] such as the cost functions for a given distance and the spatial granularity of the observation.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case. though the embedding distance outperforms geographic distance regardless of the type of gravity model used (see Supporting Information).}
\medskip

We also argue that geographical distance ﬁts well with the power-law decay function because it is is more suitable for analyzing large, complex, and scale-free spatial systems , while embedding distance ﬁts well with the exponential decay function, which we demonstrate to correspond to an underlying mathematical relationship between \textit{word2vec} and the gravity model, a connection we recently discovered and included as a major finding in this revision (see our response to Reviewer 1).

@@

\response{
\todo{I'm worry this sounds too dismissive. Edit please?}
Thank you for this comment. 
We feel that applying these embeddings to other areas of scientific activity would be incredibly enlightening! 
Your recommendation of collaboration (based on the author byline) and advisor-student pairs are especially interesting, and we would aim to address these areas in future studies. 
However, given the importance of mobility to science, the importance of human mobility more generally, and the connection we discovered between \textit{word2vec} and the gravity model, that a focus of co-affiliation, which most closely aligns with ideas of mobility, is the most reasonable data to use for this manuscript. 
}

\rcomment{
There are known concerns with clustering a 300 dimensional objects because of the curse of
dimensionality. Many have shown that reducing dimensionality with UMAP to a nontrivial
number of dimensions (e.g., 30), which preserves both local and some global distance, then
clustering on those dimensions are much more stable (e.g., this post). This was one
concern with the stability of the clustering analysis

}

Thank you for this suggestion. 
We looked into this, using UMAP to reduce the country-level embeddings from 300 down to 30 dimensions, however the results from this were incomprehensible, and inferior to those using 300 dimensions. 
Many dimensionality reduction techniques reduce the number of dimensions by maintaining pairwise distances between vectors, so that the result can be well-suited for clustering algorithms. 
However, our approach already uses raw pairwise similarities to perform hierarchical clustering, so approximating this distance in 30 dimensions, and then using the new distances to perform clustering is redundant. 
We also find that the curse of dimensional is not a major issue in our analysis, at least given the size of our data and the number of locations represented---\textit{word2vec} seems able to learn dense vector-space representations.
}

@@

\rcomment{
  My second group of questions pertain to the prestige hierarchy, that really drives the dynamics of scientific mobility. The technical description addressed most of them. However, I was wondering, is it all possible to go beyond the descriptive part, and show cases where the more accurate embedding base distances do make a difference? That is, using them, we do not only obtain better correlations with the ranking data, but also allows us to address problems that we could not without the embedding distance? For example, some results have shown that a paper coauthored by two authors at different high prestige institutions has higher impact, than writing a paper with two authors of the same high prestige institution—would embedding help us pull apart such effects? Here I am not trying to make the authors to write a different paper, so maybe a little forward thinking may be enough, if doing the actual work is too time consuming, and goes beyond the current narrative.
}

Thank you for this idea!
We believe that this is an interesting topic and we hope to investigate it in the future. 
In your particular example, an embedding distance trained using sequences of co-authorship on the same paper might be better to capture the \textit{collaborative} distance between institutions. 
Then, using that distance, one could measure the "embedding distance" spanned by all authors in the byline, and relate this to questions impact to determine whether (a)typical collaborations relate to impact. 
This is certainly a topic that is in our mind for future study. 
In the case of scientific mobility, we might also consider the "embedding distance" spanned by an individual's career, and relate this to their demographics and scientific impact.

However, while these topics are exciting, we believe that that they fall outside the scope of the current manuscript, especially given the discoveries we made concerning a fundamental connection between the gravity law of mobility and \textit{word2vec}.
}

@@

\rcomment{
Finally, one cannot escape the fact that COVID changes our mobility patterns. Many groups have used mobile phone data to explore these changes. I wonder, how would the pre-covid embeddings apply to post-covid mobility? Once again, I am mindful that this would require access to such data, but I would not be surprised at all if the authors have access to such data—in which case this is remains an intriguing question.
}

\response{
This is an excellent idea, and one that we have been interesting in addressing as well. 
Unfortunately, however, we do not have access to such data, at least not in large enough quantiaties to be representative. 
Publications are on a time lag, compared to actual events.
For example, a researcher may write a paper while studying at university $A$, but then before the final paper is published, move to university $B$, so while they now have a $B$ affiliation, in our data all we see is the $A$ one. 
As such, we would not necessarily expect to  see the impact of COVID on bibliometric data until later in 2021 or early 2022, when researchers have had the chance to publich with their new institution.
We also not that some activity will, of course, be visible, especially given the number of researchers rapidly publishing on the subject of COVID-19, however we consider this to be an unrepresentative sample, given that it is concentrated in a few disciplines.

In the future, however, we hope to have access to a fully-representative sample of mobility data to study the effects of COVID-19 on global scientific labor flows. 
We can use techniqeus developed by other researchers, such as those used by Hamilton, Leskovec, \& Jurafsky (\href{https://nlp.stanford.edu/projects/histwords/}{[link]}), to align embeddings trained on different data in order to quantify the changes in global mobility. 
We look forward to approaching this, as well as other exciting projects in the future. 
}

\clearpage
%
% Figure - Distance metric performance
%
\begin{figure}[p!]
	\centering
	\includegraphics[width=0.9\textwidth]{\figdir/Descriptive/distance_metric_performance.pdf}
	\caption{
		\textbf{Cosine distance between embedding vectors explains more flux in scientific mobility than geographic distance, baselines.}
		The correlation, or amount of flux explained by the embedding distance across distance metrics and other experimental parameters. 
		All correlations are significant at $p < 0.0001$). 
		Distance metrics are ordered, from top to bottom, by the highest correlation ($r^{2}$). 
		The asterisk denotes the top-performing distance metric for every case. 
		All distance metrics explain more of the flux than does geographic distance.
		The cosine distance between embedding vectors explains more flux than all other distance metrics. 
		The x-axis differentiates between three strategies for defining the organization population used to calculate the flux. 
		``All'' means that population is defined as the average yearly number of unique mobile and non-mobile scholars who published with the organizations' affiliation;
		population is defined in the same way for ``mobile-only'', except only using unique mobile researchers;
		``Raw frequency'' means that organization populations are defined as their frequency across all the trajectories, similar to word frequency in language embedding. 
		Generally, ``All'' explains more of the flux, however there is little difference between definitions of mobility.
		All distance metrics tend to explain more of the flux when describing all or domestic (same-country) mobility.	
		}
	\label{fig:supp:distancemetrics}
\end{figure}

### 2021-02-01 paper/current/main.tex
update rebuttal

The gravity model dictates that the expected flow, $\hat{T}_{ij}$, ($\hat{T}_{ij }= \hat{T}_{ji}$), is proportional to the locations' population, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between locations or their functional distance in the vector space, which is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}. 
A different decay function $f(r_{ij})$, which is the effect of space, can be due to fundamentally different mechanism \autocite{barthelemy2011spatial} such as the different cost function for a given distance and the spatial granularity of the observation.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:predict_emb_bymodel}), though the embedding distance outperforms geographic distance regardless of the type of gravity model used (Fig.~\ref{fig:supp:distancermse}). 
Beyond the fitting on the data, we would like to argue that geographical distance fits well with the power-law decay function because it is more suitable for analyzing large, complex, and scale-free spatial systems\autocite{chen2015distance}, while embedding distance fits well with the exponential decay function, which we will show later in the section of mathematical relation between \textit{word2vec}and the gravity model.

>>>>

The gravity model dictates that the expected flow, $\hat{T}_{ij}$, ($\hat{T}_{ij }= \hat{T}_{ji}$), is proportional to the locations' population, $\hat{T}_{ij} \propto m_{i} m_{j}$, and decays as a function of their distance, $f(r_{ij})$. 
We define the distance function in terms of either the geographic distance between locations or their functional distance in the vector space, which is calculated as the cosine distance between their vectors, termed the \textit{embedding distance}. 
The decay function $f(r_{ij})$ defines the effect of distance, and so different decay functions can model fundamentally different mechanism \autocite{barthelemy2011spatial} such as the cost functions for a given distance and the spatial granularity of the observation.
For geographic distance, we define $f(r_{ij})$ as the standard power-law function, and for the embedding distance, we use the exponential function, selected as the best performing for each case (Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:predict_emb_bymodel}), though the embedding distance outperforms geographic distance regardless of the type of gravity model used (Fig.~\ref{fig:supp:distancermse}). 
Beyond the fitting on the data, we would like to argue that geographical distance fits well with the power-law decay function because it is more suitable for analyzing large, complex, and scale-free spatial systems\autocite{chen2015distance}, while embedding distance fits well with the exponential decay function, which we will show later in the section of mathematical relation between \textit{word2vec}and the gravity model.
